### 2024-10-24 teaching/math240f24/draft.tex
notes oct 24

\medskip\boldlabel Solving congruences modulo {\mathbold n}.
Suppose we want to solve for all integers $x$ satisfying $x^2 \equiv x\pmod n$, first for $n=6$,
then for $n=7$.

@@

Now we tackle the case where $n=7$. In fact, the solution is no different
if $n$ were any other prime.
First we subtract $x$ from both sides, obtaining the congruence $x^2 - x\equiv 0\pmod 7$,
then we factorise the left side to get $x(x-1)\equiv 0\pmod 7$. Now by the previous proposition,
since $7$ is prime if $x(x-1)\equiv 0\pmod 7$, we must have either $x\equiv 0\pmod 7$ or
$x-1\equiv 0\pmod 7$, so the only solutions are $x\equiv_7 0$ and $x\equiv_7 1$ (and anything
in their equivalence classes).

@@

\proof First we prove (i). Suppose that $c_{10} c_9 c_8\cdots c_2c_1$ is a code that is
equal to $d_{10} d_9 d_8\cdots d_2 d_1$ except at one place $1\le j\le 10$. So $c_j\ne d_j$,
but for all $1\le i\le 10$ with $i\ne j$, we have $c_i = d_i$. Then
$$\eqalign{
\sum_{i=1}^{10} i c_i &\equiv_11 \sum_{i=1}^{10} i c_i - 0 \cr
&\equiv_11 \sum_{i=1}^{10} i c_i - \sum_{i=1}^{10} id_i \cr
&= \sum_{i=1}^{10} i(c_i - d_i) \cr
&= j(c_j-d_j). \cr
}$$
But $j\in \{1,\ldots, 10\}$ and $(c_j-d_j) \in [-10, 10]$. So neither $j$ nor $c_j-d_j$
are congruent to $0$ modulo $11$. Since $11$ is prime, their product cannot be congruent to $0$
modulo $11$. We conclude that $\sum_{i=1}^{10} i c_i\not\equiv 0\pmod 11$; i.e., this is
not a valid {\mc ISBN} code.

To prove part (ii), suppose that $c_{10} c_9 c_8\cdots c_2c_1$ is a code that is
equal to $d_{10} d_9 d_8\cdots d_2 d_1$, except that for some $2\le j\le 10$, we
have $c_j = d_{j-1}$ and $c_{j-1}=d_j$ (the adjacent digits are swapped). Furthermore,
assume that $d_j\ne d_{j-1}$, since swapping adjacent digits that are the same doesn't
do anything to the code. Then
$$\eqalign{
\sum_{i=1}^{10} i c_i &\equiv_11 \sum_{i=1}^{10} i c_i - 0 \cr
&\equiv_11 \sum_{i=1}^{10} i c_i - \sum_{i=1}^{10} id_i \cr
&= \sum_{i=1}^{10} i(c_i - d_i) \cr
&= (j-1)c_{j-1} + jc_j - (j-1)d_{j-1} - jd_j \cr
&= (j-1)d_j + jd_{j-1} - (j-1)d_{j-1} - jd_j \cr
&= d_{j-1} - d_j.\cr
}$$
But since $d_{j-1}$ and $d_j$ are both in the range $\{0,\ldots, 10\}$, their difference
is in the range $[-10,10]$, and we assumed that $d_{j-1}\ne d_j$, so this difference is not zero.
We conclude that $d_{j-1}-d_j\not\equiv 0\pmod 11$, so the code is not valid.\slug

@@

\medskip\boldlabel Computing large powers modulo {\mathbold p}.
The next application is not a precise mathemathical statement; rather, it's an observation
that Fermat's little theorem can allow us to compute what large powers are modulo a prime
without having to compute the number itself. We'll illustrate this by two examples.

@@

\algbegin Algorithm F (Fermat's primality test). Given an integer $n\ge 4$,
this algorithm will attempt to declare that
$n$ is not prime. If the algorithm fails to do so, then the test is inconclusive ($n$ may either
be prime or not).
\algstep F1. [Initialise.] Set $a \gets 2$. (We can skip the step $a=1$ since $a^{n-1}$ will
always equal $1$ in this case.)
\algstep F2. [Compute power.] Set $b\gets a^{n-1} \% n$. (So $0\le b< n$.)
\algstep F3. [Not prime?] If $b \neq 1$, output ``Not prime,'' and terminate the algorithm.
\algstep F4. [Loop?] Set $a\gets a+1$. If $a=n$, output ``Inconclusive,'' and terminate the algorithm.
Otherwise, go to step F2.\slug

@@

To set things up, Alice needs two large primes, call them $p$ and $q$. (Something like $200$
bits is sufficient.) Then she computes $n=pq$. Now, she picks some integer $k$ with $\gcd(k,n) = 1$.
By Theorem~{\thminvertiblemodn}, there is some integer $s$ such that $ks=1$. Alice publishes
the integer $n$ and the public key $k$. The private key is $s$; this she keeps to herself.

Now Bob wants to send a secret message $M$ to Alice. Considering $M$
as an integer (in binary), a requirement for the algorithm to work is $M<n$. Bob computes
$\bar M = M^k\% n$ and sends $\bar M$ to $A$. (We have $0\le \bar M < n$ by the division
algorithm.)

@@

Now we shall prove that the encrypted message can be decrypted into the original message.

\parenproclaim Theorem {\advthm} ({\mc RSA} encryption). Let $p$ and $q$ be distinct primes and
let $n=pq$.

>>>>

\medskip\boldlabel\datestamp{24}{x} Solving congruences modulo {\mathbold n}.
Suppose we want to solve for all integers $x$ satisfying $x^2 \equiv x\pmod n$, first for $n=6$,
then for $n=7$.

@@

Now we tackle the case where $n=7$. In fact, the solution would be no different
should $n$ be any other prime.
First we subtract $x$ from both sides, obtaining the congruence $x^2 - x\equiv 0\pmod 7$,
then we factorise the left side to get $x(x-1)\equiv 0\pmod 7$. Now by the previous proposition,
since $7$ is prime if $x(x-1)\equiv 0\pmod 7$, we must have either $x\equiv 0\pmod 7$ or
$x-1\equiv 0\pmod 7$, so the only solutions are $x\equiv_7 0$ and $x\equiv_7 1$ (and anything
in their equivalence classes).

@@

\proof First we prove (i). Suppose that $c_{10} c_9 c_8\cdots c_2c_1$ is a code that is
equal to $d_{10} d_9 d_8\cdots d_2 d_1$ except at one place $1\le j\le 10$. So $c_j\ne d_j$,
but for all $1\le i\le 10$ with $i\ne j$, we have $c_i = d_i$. Then
$$\eqalign{
\sum_{i=1}^{10} i c_i &\equiv_{11} \sum_{i=1}^{10} i c_i - 0 \cr
&\equiv_{11} \sum_{i=1}^{10} i c_i - \sum_{i=1}^{10} id_i \cr
&= \sum_{i=1}^{10} i(c_i - d_i) \cr
&= j(c_j-d_j). \cr
}$$
But $j\in \{1,\ldots, 10\}$ and $(c_j-d_j) \in [-10, 10]$. So neither $j$ nor $c_j-d_j$
are congruent to $0$ modulo $11$. Since $11$ is prime, their product cannot be congruent to $0$
modulo $11$. We conclude that $\sum_{i=1}^{10} i c_i\not\equiv 0\pmod 11$; i.e., this is
not a valid {\mc ISBN} code.

To prove part (ii), suppose that $c_{10} c_9 c_8\cdots c_2c_1$ is a code that is
equal to $d_{10} d_9 d_8\cdots d_2 d_1$, except that for some $2\le j\le 10$, we
have $c_j = d_{j-1}$ and $c_{j-1}=d_j$ (the adjacent digits are swapped). Furthermore,
assume that $d_j\ne d_{j-1}$, since swapping adjacent digits that are the same doesn't
do anything to the code. Then
$$\eqalign{
\sum_{i=1}^{10} i c_i &\equiv_{11} \sum_{i=1}^{10} i c_i - 0 \cr
&\equiv_{11} \sum_{i=1}^{10} i c_i - \sum_{i=1}^{10} id_i \cr
&= \sum_{i=1}^{10} i(c_i - d_i) \cr
&= (j-1)c_{j-1} + jc_j - (j-1)d_{j-1} - jd_j \cr
&= (j-1)d_j + jd_{j-1} - (j-1)d_{j-1} - jd_j \cr
&= d_{j-1} - d_j.\cr
}$$
But since $d_{j-1}$ and $d_j$ are both in the range $\{0,\ldots, 10\}$, their difference
is in the range $[-10,10]$, and we assumed that $d_{j-1}\ne d_j$, so this difference is not zero.
We conclude that $d_{j-1}-d_j\not\equiv 0\pmod{11}$, so the code is not valid.\slug

@@

\medskip\boldlabel Computing large powers modulo {\mathbold n}.
The next application is not a precise mathemathical statement; rather, it's an observation
that Fermat's little theorem can allow us to compute what large powers are modulo a prime
without having to compute the number itself. We'll illustrate this by two examples.

@@

\algbegin Algorithm F (Fermat's primality test). Given an integer $n\ge 4$,
this algorithm will attempt to declare that
$n$ is not prime. If the algorithm fails to do so, then the test is inconclusive ($n$ may either
be prime or not).
\algstep F1. [Initialise.] Set $a \gets 2$. (We can skip the step $a=1$ since $a^{n-1}$ will
always equal $1$ in this case.)
\algstep F2. [Compute power.] Set $b\gets a^{n-1} \percent n$. (So $0\le b< n$.)
\algstep F3. [Not prime?] If $b \neq 1$, output ``Not prime,'' and terminate the algorithm.
\algstep F4. [Loop?] Set $a\gets a+1$. If $a=n$, output ``Inconclusive,'' and terminate the algorithm.
Otherwise, go to step F2.\slug

@@

To set things up, Alice needs two large primes, call them $p$ and $q$. (Something like $200$
bits is sufficient.) Then she computes $n=pq$. Now, she picks some integer $k$ with
$$\gcd\bigl(k,(p-1)(q-1)\bigr) = 1.$$
By Theorem~{\thminvertiblemodn}, there is some integer $s$ such that $ks\equiv 1\pmod{(p-1)(q-1)}$.
Alice publishes
the integer $n$ and the public key $k$. The private key is $s$; this she keeps to herself.

Now Bob wants to send a secret message $M$ to Alice. Considering $M$
as an integer (in binary), a requirement for the algorithm to work is $M<n$. Bob computes
$\bar M = M^k\percent n$ and sends $\bar M$ to $A$. (We have $0\le \bar M < n$ by the division
algorithm.)

@@

Now we shall prove that the encrypted message $\bar M = M^k$
can be decrypted into the original message.

\parenproclaim Theorem {\advthm} ({\mc RSA} encryption). Let $p$ and $q$ be distinct primes and
let $n=pq$. Let $k$ and $s$ be such that $ks\equiv 1\pmod n$. Then for all integers $0\le M < n$,
$(M^k)^s \equiv M\pmod n$.

\proof Without loss of generality, we can take $k$ and $s$ both positive. From
$ks\equiv 1\pmod{(p-1)(q-1)}$,
we know there is an integer $l$ such that $ks-1 =l (p-1)(q-1)$. Since $k,s>0$, this identity
means that $l>0$ as well. Now, since $n=pq$, we have
$$(M^k)^s \equiv M^{ks} \equiv M^{1+l(p-1)(q-1)} \equiv M(M^{p-1})^{l(q-1)} \pmod{pq}.$$
By part (i) of the previous lemma, this means we have the same congruence modulo $p$; i.e.,
$$(M^k)^s \equiv M(M^{p-1})^{l(q-1)} \pmod p,$$
and by Fermat's little theorem $M^{p-1}\equiv 1$ for all $M\not\equiv 0\pmod p$, so this simplifies to
$$\eqalign{
(M^k)^s &\equiv_p \cases{M, & if $M\not\equiv 0\pmod p$; \cr 0, & otherwise} \cr
&\equiv_p M.\cr
}$$
{\it Mutatis mutandis} (by the roles of $p$ and $q$ in the above), we have
$(M^k)^s \equiv_q M$ as well, so part (ii) of the previous lemma can now be applied to give
$$(M^k)^s \equiv M \pmod{pq}.\noskipslug$$

Here is an example of the {\mc RSA} algorithm in action. Alice sets $p=3$ and $q=11$,
so that $n=33$. This means that $(p-1)(q-1) = 2\cdot 10 = 20$. Using $k=7$ (which we can
do since $\gcd(7,20) = 1$, we have $s=3$, since $3\cdot 7 = 21\equiv 1\pmod{20}$.
Alice publishes the integer $n=33$ as well as the public key $k=7$. She keeps the number $s=3$ private.

The possible messages $M$ one can send to Alice are $0,1,\ldots, 32$, since we need $0\le M<n$.
Suppose Bob wants to send the message $00010$ to Alice; that is, we have $M=2$. Bob computes
$$M^k = 2^7 = 2^5\cdot 2^2 = 32\cdot 4 \equiv (-1)\cdot 4 \equiv 29\pmod{33},$$
and sends Alice $\bar M = 29$.

Now Alice computes
$$\bar M^s = 29^3 \equiv (-4)^3 = -64 \equiv -64 + 66 = 2\pmod{33}$$
and recovers the original message $M=2$.

Now imagine that there is a third person, Charlie, who wants to thwart Alice and Bob.
The integer $n$ and the public key $n$ are known to all,
including Charlie. If he wants to decrypt Bob's message, all Charlie needs to know is the
the private key $s$. In this case, since $n$ is very small, the Adversary can simply
factor $33 = 3\cdot 11$ to find out that $(p-1)(q-1) = 20$, and then use Euclid's Algorithm
to find the inverse of $7$ modulo $20$, and he will be able to decrypt all messages between
Alice and Bob.

In real life, {\mc RSA} is implemented using very large primes $p$ and $q$.
This makes it difficult, but not impossible, to break the encryption. To put things
in perspective, it took a team of researchers $900$ core-years of computing time to
factorise the $240$-digit ($795$-bit) modulus $n$ that is used
 in {\mc RSA}-240 encryption. These researchers estimated
that increasing the number of bits to $1024$ would make their factorisation program
take $500$ times as long to crack the encryption.

However, this does not mean there doesn't exist a fast algorithm for factorising
integers. We just haven't found one yet (or if people have, they are keeping the algorithm
to themselves). Most computer scientists believe that this is impossible though. It
is conjectured
that integer factorisation is a difficult problem---that is, there is no algorithm to
factorise a $b$-bit integer in time polynomial in $b$. Until this conjecture is proved,
all of the data we rely on encryption to protect could theoretically be in jeopardy.

\vfill\eject
\begingroup\headline{\hfil}\footline{\hfil}
\vglue160pt
\centerline{\titlefont III. GRAPH THEORY}
\vskip220pt
\bigskip
\begingroup\obeylines\eightssi
\hfill Quamobrem, cum nuper problematis cuiusdam mentio esset facta,
\hfill quod quidem ad geometriam pertinere videbatur,
\hfill at ita erat comparatum, ut neque determinationem quantitatum requieret,
\hfill neque solutionem calculi quantitatum ope admitteret,
\hfill id ad geometriam situs referre haud dubitavi:
\hfill praesertim quod in eius solutione solus situs in considerationem veniat,
\hfill calculus vero nullius prorsus sit usus.
\eightss
\smallskip
\hfill --- LEONHARD EULER {\eightssi ``Solutio problematis ad geometriam situs pertinentis''} (1741)
\endgroup%\obeylines
\bigskip\goodbreak
\vfill\eject

\advsect Definitions and basic notions

A {\it graph} is a pair $G = (V,E)$ where $V$ is a nonempty set and
$$E \subseteq \bigl\{ \{u,v\} : u,v\in V\bigr\}.$$
The elements of $V$ are called {\it vertices} and the elements of $E$ are
called edges.
If $e = \{u,v\}\in E$, we say that $u$ and $v$ are {\it adjacent}, and
$e$ is {\it incident on} $u$ and $v$. For brevity, we often write $uv$ instead
of $\{u,v\}$.

For instance, we might have $V = \{1,2,3,4\}$ and
$$E = \{12, 13, 14, 24\}.$$
We can draw the graph $G = (V,E)$ as follows:
\midinsert
$$\epsfbox{firstexample.ps}$$
\endinsert
\goodbreak
More examples of graphs are pictured in the figure below.

\midinsert
$$\epsfbox{smallgraphs.ps}$$
\vskip5pt
\caption{Some small graphs.}
\endinsert
\goodbreak

We can define graphs in a more complicated manner. Let $G = (V,E)$
be the graph with $V = \ZZ/5\ZZ$ and $ab\in E$ if and only if
$$a + b \equiv 1\pmod 5\qquad\hbox{or}\qquad a+b \equiv 3\pmod 5.$$
Here is a drawing of $G$:
\midinsert
$$\epsfbox{modfiveexample.ps}$$
\endinsert
\goodbreak
